Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-26 inference.tex
GKS decidability.

    & & \hspace*{1.45in}}

@@

    & & \hspace*{1.45in}} {\bot}) \Rightarrow g_0(\C{item\_req},s')\\

@@

    & & \hspace*{0.2in}\forall \C{ireq}. \forall s.~ 
        \pi_3 \Leftrightarrow (\C{ireq} \in \C{item\_reqs} \conj 
        (\G_U(\C{ireq},s') \vee \G_I(\C{ireq},s'))

@@

    & & \conj \hspace*{0.05in}\forall s.\exists \C{ireq}. ~ 
        \pi_4 \Leftrightarrow (g_1(s) \Rightarrow (\C{ireq} \in \C{item\_reqs} 
        \conj (\G_U(\C{ireq},s') \vee \G_I(\C{ireq},s')))\\

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a subset of logic that satisfies the following syntactic properties:
\begin{itemize}
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item All quantifiers are universal, and appear only at the prenex
    position, i.e., at the beginning of a quantified formula.  
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
\end{itemize}
This fragment of FOL known as EPR (Effectively Propositional logic) is
known to be decidable~\cite{z3epr}. The language of encoding, however,
is a combination of {\sf EPR} with (a). $\Prop^0$, the theory from
which quantifier-free propositions ($\varphi$) that encode object
language expressions are drawn, and (b).  $\Prop^1$, the theory from
which invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
and state the following theorem:
\begin{theorem}
  $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
\begin{theorem}
  $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
\end{theorem}
The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
describe common database integrity constraints, such as referential
integrity, non-negativeness of all integer values in a column etc.
The isolation specifications presented in \S\ref{sec:isolation} are
already simple first-order formulas that can be encoded in {\sf EPR}.
Furthermore, it is also reasonable to expect the guarantee ($G$) of a
transaction to be expressible in the same logic as its inferred $\F$,
since $\F$ (without the stability check) is essentially a complete
characterization of the transaction, while $G$ is only an abstraction.
Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
inference, the verification problem for weakly isolated transactions
is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
making automated verification a practical exercise.

>>>>

    & & \hspace*{0.85in}}

@@

    & & \hspace*{0.85in}} {\bot}) \Rightarrow g_0(\C{item\_req},s')\\

@@

    & & \hspace*{0.2in}\forall \C{item\_req}. \forall s.~ 
        \pi_3 \Leftrightarrow (\C{item\_req} \in \C{item\_reqs} \conj 
        (\G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s'))

@@

    & & \conj \hspace*{0.05in}\forall s.\exists \C{item\_req}. ~ 
        \pi_4 \Leftrightarrow (g_1(s) \Rightarrow (\C{item\_req} \in \C{item\_reqs} 
        \conj (\G_U(\C{item\_req},s') \vee \G_I(\C{item\_req},s')))\\

@@

Observe that the encoding shown in Fig.~\ref{fig:logic} maps to
a fragment of FOL that satisfies the following syntactic properties:
\begin{itemize}
  \item All function symbols, modulo those that might appear in
    $\varphi\in\Prop^0$, are uninterpreted and Boolean.
  \item All quantification is first-order; second-order objects, such
    as sets and functions, are never quantified.
  \item Quantifiers appear only at the prenex position, i.e., at the
    beginning of a quantified formula.
\end{itemize}
The simple syntactic structure of the fragment already makes is
amenable for automatic reasoning via an off-the-shelf SMT solver, such
as Z3. The decidability of this fragment, however, is more subtle and
discussed below.

Consider a set expression $s$ with no free variables (i.e., $\nubar =
\emptyset$, like $\T(\stl,\stg)$ from the above example). Let
$(\phi,G) = \mssemof{\emptyset}{s}$. Note that $\phi$ is a conjunction
of (a).  $\phi_i$'s, where each $\phi_i$ results from encoding a
subexpression $s_i$ of $s$, and (b). a $\phi_s$, resulting from
encoding $s$ itself (i.e., its top-level expression). From Fig.~\ref{fig:logic},
it is clear that $\phi_s$ is either $\top$ (for the first four cases),
or it is a prenex-quantified formula, where quantification is either
$\forall^2$, or $\exists$, or $\forall\exists$. Generalizing this
observation, for a set expression $s$ with $|\nubar|$ free variables,
$\phi_s$, if quantified, is a prenex-quantified formula, where
quantification assumes one among the forms of $\forall^{|\nubar|+2}$,
or $\forall^{|\nubar|}\exists$, or $\forall^{|\nubar|+1}\exists$.  In
other words, the number of $\forall$ quantifiers preceding an
$\exists$ quantifier is utmost two more than the number of free
variables ($\nubar$) in $s$. For the convenience of this discussion,
let us call $\forall^{|\nubar|+1}\exists$ as the prenex signature of
$\phi_s$.

Next, in Fig.~\ref{fig:logic}, observe that the (ordered) set $\nubar$
is extended only in the encoding rules for $\bind$ and $\existsl$
expression. Since $\existsl$ is a result of a particular choice of the
weakening semantics (Sec.~\ref{sec:inference}), let us ignore it to
focus on the core set expression language. This leaves $\bind$ as the
only case where $\nubar$ is extended.  Since an occurance of $\bind$
adds one to $|\nubar|$, if $s$ is a subexpression of a top-level
$\bind$ expression (like $\F_U(\stl,\stg)$ from the above example),
then the prenex signature of $\phi_s$ is $\forall^2\exists$.
Furthermore, if the subexpressions of $s$ are not $\bind$ expressions,
then none of the $\phi_i$'s are quantified, and the prenex signature
of $\phi = \bigwedge_i\phi_i \wedge \phi_s$ remains
$\forall^2\exists$.  The fragment of FOL that admits formulas with
prenex signatures of form $\forall^2\exists^*$ is called
Godel-Kalmar-Schutte ({\sf GKS}) fragment, which is known to be
decidable. The language of encoding, however, is a combination of {\sf
GKS} with (a). $\Prop^0$, the theory from which quantifier-free
propositions ($\varphi$) that encode object language expressions are
drawn, and (b).  $\Prop^1$, the theory from which invariants ($I$) are
drawn. Thus the encoding is decidable if the combination of ${\sf GKS}
+ \Prop^0 + \Prop^1$ is decidable.

To formally state the above property, let $\SL[n,\Prop^0,\Prop^1]$
denote the core set expression language that admits $n$ nested $\bind$
expressions, and allows program's (boolean) expressions ($e$), and
invariants ($\I$) to be drawn from theories $\Prop^0$ and $\Prop^1$,
respectively. Then, we have the following result:
\begin{theorem}
  $\SL[0,\Prop^0,\Prop^1]$ is decidable if ${\sf GKS}+\Prop^0+\Prop^1$
  is decidable.
\end{theorem}
The $n=0$ case denotes the lack of nested bind expressions. Since
nested \C{foreach} statements are the only program constructs in
$\txnimp$ that lead to nested bind expressions, the $n=0$ case
essentially denotes the encoding language for $\txnimp$ programs
without nested \C{foreach} loops. Observe that the \C{new\_order}
transaction (Fig.~\ref{fig:new_order_code}) satisfies this constraint.
Indeed, the restricted language without nested \C{foreach} loops is
nonetheless expressive enough to cover all TPC-C transactions.

% This
% fragment of FOL known as EPR (Effectively Propositional
% logic) is
% known to be decidable~\cite{z3epr}. The language of encoding, however,
% is a combination of {\sf EPR} with (a). $\Prop^0$, the theory from
% which quantifier-free propositions ($\varphi$) that encode object
% language expressions are drawn, and (b).  $\Prop^1$, the theory from
% which invariants ($I$) are drawn. We write $\SL[\Prop^0,\Prop^1]$ to
% highlight the parameterization of $\SL$ on $\Prop^0$ and $\Prop^1$,
% and state the following theorem:
% \begin{theorem}
%   $\SL[\Prop^0,\Prop^1]$ is decidable if ${\sf EPR}+\Prop^0+\Prop^1$
%   is decidable.
% \end{theorem}
% A useful instantiation of $\SL$ is $\SL[{\sf SLA},{\sf EPR}+{\sf
% SLA}]$, where ${\sf SLA}$ is the theory of simple linear arithmetic.
% Since {\sf EPR}+{\sf SLA} is known to be decidable~\cite{eprsla}:
% \begin{theorem}
%   $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ is decidable.
% \end{theorem}
% The $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ instantiation requires $I$ to
% be drawn from {\sf EPR}+{\sf SLA}, which is expressive enough to
% describe common database integrity constraints, such as referential
% integrity, non-negativeness of all integer values in a column etc.
% The isolation specifications presented in \S\ref{sec:isolation} are
% already simple first-order formulas that can be encoded in {\sf EPR}.
% Furthermore, it is also reasonable to expect the guarantee ($G$) of a
% transaction to be expressible in the same logic as its inferred $\F$,
% since $\F$ (without the stability check) is essentially a complete
% characterization of the transaction, while $G$ is only an abstraction.
% Thus, with $\SL[{\sf SLA},{\sf EPR}+{\sf SLA}]$ as the language of
% inference, the verification problem for weakly isolated transactions
% is decidable. Moreover, off-the-shelf SMT solvers (e.g., Z3) are
% equipped with efficient decision procedures for ${\sf EPR}+{\sf SLA}$,
% making automated verification a practical exercise.
